---
title: "Alternative TAP-seq Predictive Models"
author: "Micah Olivas"
date: "11/11/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = F, message = F)
options(bitmapType = "cairo")
```

## Compute environment

Clear data environment. 

```{r ClearDataEnv}
rm(list=ls())
```

Load required packages and functions.

```{r setupPackages, message=F, warning=F}
.libPaths("/g/steinmetz/velten/Software/RLibs-seurat3/")
library("tidyverse")
library("randomForest")
library("caret")
library("dplyr")
library("ranger")
library("xgboost")
library("MLeval")
library("MLmetrics")
library("PRROC") # for Precision-Recall curve calculations
library("ROCR")
```


## Data

Differential expression testing was performed for the data from the screen and results were annotated with epigenome and HiC information. To generate these data, run the rule the rule chromatin_annotated_etps of https://github.com/argschwind/tapseq_manuscript. The results are loaded into R. Gene-wise mean expression levels are also loaded.

```{r loadOriginalData}
pairs_old = read.csv(url("http://steinmetzlab.embl.de/TAPdata/chromatin_annotated_pairs.csv"), stringsAsFactors = F)
#this corresponds to the ../data/chromatin_annotated_pairs.csv if the snakemake pipeline was run
expression = readRDS(url("http://steinmetzlab.embl.de/TAPdata/screen_mean_expression.RDS"))
#this corresponds to the row means of ../data/11iScreen1 if the snakemake pipeline was run
```

```{r loadFilteredData}
pairs = read.csv("/Users/micaholivas/Desktop/Rotations/Steinmetz_Fall/Filtered_Data/chrom_tapseq_perCRE_nGenesCovar_0.08gStd.tsv", stringsAsFactors = F)
#this corresponds to the ../data/chromatin_annotated_pairs.csv if the snakemake pipeline was run
expression = readRDS(url("http://steinmetzlab.embl.de/TAPdata/screen_mean_expression.RDS"))
#this corresponds to the row means of ../data/11iScreen1 if the snakemake pipeline was run

pairs <- pairs %>%
  dplyr::filter(significant == 1 | PowerAtEffectSize25 >= 0.8) %>%
  dplyr::select(-c(PowerAtEffectSize10, PowerAtEffectSize25, PowerAtEffectSize50))
```

For comparing classifier performance between datasets, we annotate the enhancers used by [Gasperini et al.](https://www.ncbi.nlm.nih.gov/pubmed/30612741) in the same way and load it into R. Also, we load their differential expression testing results (Table S2b of Gasperini et al.)

```{r loadGasperini}
# gasperini.predictors = read.csv(url("http://steinmetzlab.embl.de/TAPdata/chromatin_annotated_pairs_gasperini_screen.csv"), stringsAsFactors = F)
# gasperini.hits = read.csv(url("http://steinmetzlab.embl.de/TAPdata/gasperini_hits_screen.csv"),sep=";")

```


## Global Data Preparation

```{r prepareData}
pairs$random = rnorm(nrow(pairs)) # add a random covariate
pairs = subset(pairs, grepl("^chr", perturbation)) # removes controls
pairs$expression = expression[as.character(pairs$gene)] # add baseline expression level of gene
pairs$enh_chr = factor(pairs$enh_chr) # Select genes which are associated with at least one enhancer within 300kb

sig_genes_pairs_300kb_atLeastOneHit = pairs %>%
  group_by(sample, gene) %>%
  filter(dist_to_tss <= 0.3e6 ) %>% # changed, like in fulco
  filter(sum(significant) > 0) %>% 
  select(-c(2, 5:13, 24:28, 31))
  
# Genes which are associated with at least one enhancer genome-wide
sig_genes_pairs = pairs %>%
  group_by(sample, gene) %>%
  filter(sum(significant) > 0) %>%
  select(-c(2, 5:13, 24:28, 31))

```


## Partition data for Cross-validation

```{r CV_Partition}

# ABC_norm = sig_genes_pairs_300kb_atLeastOneHit$abc_score_norm
# dist_bins = c("dist_bins")
# ABC_score_dist_bins_expression = c(ncol("abc_score_norm", "abc_score_norm", "dist_bins", "expression"))

# Partition data for cross-validation during scripting
inTraining = createDataPartition(sig_genes_pairs_300kb_atLeastOneHit$significant, p = .90, list = FALSE)
training = sig_genes_pairs_300kb_atLeastOneHit[ inTraining, -c(14,16,17,19)] # Remove ABC score, dist_bins
testing = sig_genes_pairs_300kb_atLeastOneHit[-inTraining, -c(14,16,17,19)] # Remove ABC score, dist_bins


chrom8_sig_genes_pairs_300kb_atLeastOneHit = sig_genes_pairs_300kb_atLeastOneHit %>% 
  subset(subset = sig_genes_pairs_300kb_atLeastOneHit$enh_chr == "chr8")

chrom11_sig_genes_pairs_300kb_atLeastOneHit = sig_genes_pairs_300kb_atLeastOneHit %>% 
  subset(subset = sig_genes_pairs_300kb_atLeastOneHit$enh_chr == "chr11")


chrom8_training = training %>% 
  subset(subset = training$enh_chr == "chr8")

chrom11_training = training %>% 
  subset(subset = training$enh_chr == "chr11")

chrom8_testing = testing %>% 
  subset(subset = testing$enh_chr == "chr8")

chrom11_testing = testing %>% 
  subset(subset = testing$enh_chr == "chr11")


preproc_list = list(
  model_combined_training = training, #  Change data set name and contents here
  model_combined_testing = testing,
  model_chrom8_training = chrom8_training,
  model_chrom8_testing = chrom8_testing,
  model_chrom11_training = chrom11_training,
  model_chrom11_testing = chrom11_testing
  # training_gw = training_gw, 
  # testing_gw = testing_gw
  )


# Create function to remove non-numeric data testing, training datasets
process_data = function(my_data){
  numvar = names(my_data)[which(sapply(my_data, is.numeric))]
  my_data_new <<- my_data[,numvar]
  
  my_data_new$significant <<- ifelse(my_data_new$significant == 1, "yes", "no")
  my_data_new$significant <<- factor(my_data_new$significant) # factorize significaance column

  return(my_data_new)
}

# Apply numeric selction function to list of datasets 
output = lapply(preproc_list, process_data)
for(i in 1:length(output)){
  tempobj = output[[i]]
  eval(parse(text = paste(names(output)[[i]], "=tempobj")))
}


```


## Configure Caret Training Controls

```{r Training_Controls}

# Set training controls
fitControl = trainControl(method = "repeatedcv",
                           number = 10,
                           repeats = 10,
                           ## Estimate class probabilities
                           classProbs = TRUE,
                           savePredictions = TRUE,
                           ## Evaluate performance using the following function
                           summaryFunction = prSummary
                           )

```


## Train Combined Pairs models in Caret

```{r Combined_Train}


# Train model with Caret
fit_combined_ranger = train(significant ~ .,
                            data = model_combined_training,
                            method = "ranger",
                            trControl = fitControl,
                            verbose = FALSE,
                            ## Specify which metric to optimize
                            metric = "AUC"
                            )

# Train GBT model
fit_combined_xgbt = train(significant ~ .,
                          data = model_combined_training,
                          method = "xgbTree",
                          verbose = FALSE,
                          trControl = fitControl,
                          ## Specify which metric to optimize
                          metric = "AUC"
                          )

# Train GBM model
fit_combined_GBM = train(significant ~ .,
                         data = model_combined_training,
                         method = "gbm",
                         verbose = FALSE,
                         trControl = fitControl,
                         ## Specify which metric to optimize
                         metric = "AUC"
                         )

```


## Train on Chromosome 8 pairs

```{r Chrom8_Train}

# Train rf model with Caret
fit_chrom8_ranger = train(significant ~ .,
                  data = model_chrom8_training,
                  method = "ranger",
                  trControl = fitControl,
                  verbose = FALSE,
                  ## Specify which metric to optimize
                  metric = "AUC"
                  )

# Train GBT model
fit_chrom8_xgbt = train(significant ~ .,
                  data = model_chrom8_training,
                  method = "xgbTree",
                  trControl = fitControl,
                  verbose = FALSE,
                  ## Specify which metric to optimize
                  metric = "AUC"
                  )

# Train GBM model
fit_chrom8_GBM = train(significant ~ .,
                  data = model_chrom8_training,
                  method = "gbm",
                  trControl = fitControl,
                  verbose = FALSE,
                  ## Specify which metric to optimize
                  metric = "AUC"
                  )

```


## Train on Chromosome 11 Pairs

```{r Chrom11_Train}

# Train model with Caret
fit_chrom11_ranger = train(significant ~ .,
                  data = model_chrom11_training,
                  method = "ranger",
                  trControl = fitControl,
                  verbose = FALSE,
                  ## Specify which metric to optimize
                  metric = "AUC"
                  )

# Train xGBT model
fit_chrom11_xgbt = train(significant ~ .,
                  data = model_chrom11_training,
                  method = "xgbTree",
                  trControl = fitControl,
                  verbose = FALSE,
                  ## Specify which metric to optimize
                  metric = "AUC"
                  )

# Train GBM model
fit_chrom11_GBM = train(significant ~ .,
                  data = model_chrom11_training,
                  method = "gbm",
                  trControl = fitControl,
                  verbose = FALSE,
                  ## Specify which metric to optimize
                  metric = "AUC"
                  )

```


## Self Comparisons

Calculate AUPRC for Self Models
```{r Compare_Models}

# Create a list containing all models
list_combined_models = list(xgbt_combined = fit_combined_xgbt,
                            GBM_combined = fit_combined_GBM,
                            random_forest_combined = fit_combined_ranger
                            )

list_chrom8_self_models = list(xgbt_chrom8 = fit_chrom8_xgbt,
                               GBM_chrom8 = fit_chrom8_GBM,
                               random_forest_chrom8 = fit_chrom8_ranger
                               )

list_chrom11_self_models = list(xgbt_chrom11 = fit_chrom11_xgbt,
                                GBM_chrom11 = fit_chrom11_GBM,
                                random_forest_chrom11 = fit_chrom11_ranger
                                )


# Write function to generate predictions and precision recall curve by model on testing set
calc_auprc = function(model, data){
  index_significant = data$significant == "yes"
  index_insignificant = data$significant == "no"
  
  predictions = predict(model, data, type = "prob")
  
  pr.curve(predictions$yes[index_significant], predictions$yes[index_insignificant], curve = TRUE)
}

# calc_auprc(fit_chrom8_GBM, model_chrom11_training)

# mat = cbind(predict(fit_combined_ranger, model_combined_training, type = "prob"),
#             model_combined_training$significant)
# 
# model_combined_training$significant[1:100]
# 
# combined_preds = predict(fit_combined_ranger, model_combined_training, type = "vote")
# combined_preds[1:100]
# view(fit_combined_GBM$pred)
# 
# y <- as.data.frame(pr$curve)
# ggplot(y, aes(y$V1, y$V2))+geom_path()+ylim(0,1)




# Calculate AUPRC for models
# PR_combined_preds = list_combined_models %>%
#   map(calc_auprc, data = model_combined_testing)
# 
# AUPRCs = c(PR_combined_preds$xgbt_combined$auc.integral,
#            PR_combined_preds$GBM_combined$auc.integral,
#            PR_combined_preds$random_forest_combined$auc.integral,
#            0.5364040 # From Schraivogel et al. 2020 combined training script
#            )

model_types = c("XGBOOST", "GBM", "Random Forest (Caret)", "Random Forest (Old)")
metrics = c("Train on All")

model_metrics = list(metrics, model_types)

AUPRC_summaries = matrix(AUPRCs, ncol = length(model_types), dimnames = model_metrics)
```


Plot AUPRC for Combined and Self Chrom Prediction Training
```{r Plot_Self_Curves}

# Plot Curves for Combined Train and Test
curves_combined = evalm(list(fit_combined_ranger, fit_combined_xgbt, fit_combined_GBM), gnames = c("Random Forest", "Gradient-Boosted Trees", "Gradient-Boosted Machine"), plots = c("pr"), title = "AUPRCs (Trained on Significant Pairs less than 300kb separation)")

```


## Non-self Comparisons

Calculate area under PR curves for non-self prediction
```{r Other_AUPRC}

# Calculate AUPRC for models
PR_8on11_preds = list_chrom8_self_models %>%
  map(calc_auprc, data = model_chrom11_testing)

PR_11on8_preds = list_chrom11_self_models %>%
  map(calc_auprc, data = model_chrom8_testing)

AUPRCs_8on11 = c(PR_8on11_preds$xgbt_chrom8$auc.integral,
                 PR_8on11_preds$GBM_chrom8$auc.integral,
                 PR_8on11_preds$random_forest_chrom8$auc.integral,
                 0.4083787 # From Schraivogel et al. 2020 combined training script
                 )

AUPRCs_11on8 = c(PR_11on8_preds$xgbt_chrom11$auc.integral,
                 PR_11on8_preds$GBM_chrom11$auc.integral,
                 PR_11on8_preds$random_forest_chrom11$auc.integral,
                 0.5122513 # From Schraivogel et al. 2020 combined training script
                 )

PR_11on8_preds

model_types = c("GBM", "XGBOOST", "Random Forest (Caret)", "Random Forest (Old)")
metrics_8on11 = c("8 -> 11")
metrics_11on8 = c("11 -> 8")

model_metrics_8on11 = list(metrics_8on11, model_types)
model_metrics_11on8 = list(metrics_11on8, model_types)

AUPRC_summaries = rbind(AUPRC_summaries, matrix(AUPRCs_8on11, ncol = length(model_types), dimnames = model_metrics_8on11))
AUPRC_summaries = rbind(AUPRC_summaries, matrix(AUPRCs_11on8, ncol = length(model_types), dimnames = model_metrics_11on8))

```

Plot PR curves for non-self prediction 
```{r Plot_Other_Curves}


# list_chrom8on11_models = list(xgbt_chrom8on11 = fit_chrom8on11_xgbt,
#                           GBM_chrom8on11 = fit_chrom8on11_GBM,
#                           random_forest_chrom8on11 = fit_chrom8on11_ranger
#                           )
# 
# list_chrom11on8_models = list(xgbt_chrom11on8 = fit_chrom11on8_xgbt,
#                            GBM_chrom11on8 = fit_chrom11on8_GBM,
#                            random_forest_chrom11on8 = fit_chrom11on8_ranger
                           # )

# Map receiver operating characteristic curve function to lists of single models
PR_8on11_preds = list_chrom8_self_models %>%
  map(calc_auprc, data = model_chrom11_testing)

PR_11on8_preds = list_chrom11_self_models %>%
  map(calc_auprc, data = model_chrom8_testing)

# Plot Curves for Chrom 8 to Chrom 11
curves_8on11 = evalm(list(fit_chrom8_ranger, fit_chrom8_xgbt, fit_chrom8_GBM), gnames = c("Random Forest", "Gradient-Boosted Trees", "Gradient-Boosted Machine"), plots = c("pr"), title = "AUPRC Comparison of Combined on 11")

# Plot Curves for Chrom 8 to Chrom 11
curves_11on8 = evalm(list(fit_chrom11_ranger, fit_chrom11_xgbt, fit_chrom11_GBM), gnames = c("Random Forest", "Gradient-Boosted Trees", "Gradient-Boosted Machine"), plots = c("pr"), title = "AUPRC Comparison of Combined on 11")


```


Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
